Relationships in the extremes and their influence on competition and coexistence

### 2021-09-03 paper/Paper.Rnw
edits, done for now with Methods and Results

%G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
%from ATAs compared to other mechanisms. This is to help illuminate whether ATA contributions to coexistence 
%are likely to be negligible or meaningfully large.

Left- or right-tail associated noise $(b_1,b_2)$, with $b_i$ normally distributed with mean $\mu_i$ and standard deviation 
$\sigma$, was generated as pictured in Fig. \ref{fig:noise}, with details in SI section \ref{SI-sect:noise}.
To achieve goal G2 from the Introduction, i.e., to explore using a simple model the relative magnitudes of the contributions 
to coexistence that come from ATA versus other mechanisms, we plotted the terms $\Delta_i^0$, $\Delta_i^E$, $\Delta_i^C$, 
$\Delta_i^{(E\shp C)}$, $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ [see (\ref{eq:full_decomp})] for the lottery model
% and $r_1-q_{12}r_2$?
against the independent parameters of the model, $\mu_1$, $\mu_2$, $\sigma$ and $\delta$; and then we compared their
values on the plots. Some mathematical shortcuts, summarized here, greatly simplified this plotting. First, we showed that, for the 
lottery model, none of the components in (\ref{eq:full_decomp}) depends independently on $\mu_1$ and $\mu_2$;
instead they all depend only on the difference $\mu_1-\mu_2$ (SI section \ref{SI-sect:math_simplifications}). So we only plotted 
against $\mu_1 - \mu_2$. Second, we showed that the values of all components in (\ref{eq:full_decomp}) were the same 
for both our left- and right-tail associated noises (SI section \ref{SI-sect:math_simplifications}). So we only generated plots 
for left-tail associated noise. For each combination of $\delta=0.2, 0.4, 0.6, 0.8$
and $\mu_1-\mu_2=0,-0.5,-2,-4$,
we plotted all the components of (\ref{eq:full_decomp}) against $\sigma$. 
Relative importance of contributions was judged from the 
plots to assess whether and to what extent ATA effects on coexistence were meaningfully large for the lottery model, compared to 
the contributions of other mechanisms.
Mathematical simplications for the $\delta=1$ case make it straightforward to see that
coexistence is impossible and that the case is relatively trivial (SI 
section \ref{SI-sect:delta1}). So that case is not considered further.  
The simulation methods of \cite{Ellner_2019} were used to compute the components of (\ref{eq:full_decomp}).
Computationally efficient means of carrying out these simulations, for the lottery model, are elaborated in 
SI section \ref{SI-sect:efficient_computation}.

%G3) To explore, using our simple model, the likelihood that species can coexist with
%ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions biodiversity maintenance 
%may be negligible or meaninfully large.

To acheive goal G3 from the Introduction i.e., to explore, using the lottery model, the likelihood that species can coexist with
ATAs but not without, or vice versa, we plotted the IGR, $\olr_{i \bs i}=\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$, 
as well as the
IGR minus its ATA contribution, $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$. 
For fixed values of $\sigma$ and $\delta$ and for left-tail associated noise, 
we were interested in the values of $\mu_1-\mu_2$ for which one of these
quantities is positive and the other is negative; this set of value characterizes the species which can coexist
with ATAs, but not without, or vice versa. The one dimensional space of values $\mu$ is a species trait space,
and $\mu_1$ and $\mu_2$ can be regarded as two points in it. But we instead define the ``trait-difference space'' of values $\mu_1-\mu_2$. 
Points of the trait-difference space characterize possible combinations of traits of the two competing species in our model. For fixed values of 
$\delta$ and $\sigma$, we defined the subset, $\psi_{\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of 
values $\mu_1 - \mu_2$ for which the the IGR, $\olr_{i \bs i}$, is positive. This 
is the subspace for which the two species in the lottery model coexist in the presence of ATAs.
We also defined the subset, $\psi_{\lnot\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of values
$\mu_1 - \mu_2$ for which the metric $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$ is positive.
This is the subspace for which the two species in the lottery model coexist in the absence of 
ATAs. The symbol $\lnot$ denotes logical negation, so $\lnot \text{ATA}$ is intended to indicate the
absence of ATAs. We defined the ``ATA rescue space" associated with given values of $\delta$ and $\sigma$ to be the 
set difference $\psi_{\text{ATA}}(\delta,\sigma) \setminus \psi_{\lnot\text{ATA}}(\delta,\sigma)$, i.e., the
values $\mu_1 - \mu_2$ for which species coexist in the presence of ATAs, but not in their absence. 
Similarly, we define ``ATA exclusion space'' to be the set difference 
$\psi_{\lnot\text{ATA}}(\delta,\sigma) \setminus \psi_{\text{ATA}}(\delta,\sigma)$,
i.e., $\mu_1 - \mu_2$ values for which coexistence occurs in the absence, but not the presence, of ATAs.
On plots of $\olr_{i \bs i}$ and $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$
against $\mu_1 - \mu_2$, we highlighted the ATA rescue and exclusion spaces. We plotted against
$\mu_1 - \mu_2$ for all combinations of $\delta=0.2, 0.4, 0.6, 0.8$ and $\sigma=2,4,5,6$.

@@

%***DAN: Some version of the below text ought to go in the figure caption, do that later, Dan
%The spaces in between the lines where they intersect the x-axis at y=0, indicate either an ATA rescue trait space or an ATA exclusion trait space. If the line plotting IGR is to the left of the line plotting IGR without ATA, then this space is an ATA rescue trait space. If the line plotting IGR is to the right of the line plotting IGR without ATA, then this space is an ATA exclusion trait space.

@@

%G2 and related side results
Fulfilling goal G2 of the Introduction (to explore, using the lottery model, the relative magnitudes 
of the contributions to coexistence that come from ATAs compared to other mechanisms), Fig.
\ref{fig:fig3} shows that the ATA term of (\ref{eq:full_decomp}) sometimes, but not always, has 
magnitude comparable to the magnitudes of other terms of the extended MTC decomposition. 
Magnitudes of $\Delta_i^{[E \nat C]}$ tended to be larger, relative to the magnitudes of the 
other terms of (\ref{eq:full_decomp}), for smaller-magnitude differences between $\mu_1$ and 
$\mu_2$, i.e., for closer-to-equal competitors. 
In the $\mu_1=\mu_2$ case (panels A-D), for which the two species were equal 
competitors, storage effects (blue plus red
lines on Fig. \ref{fig:fig3}) were the only means by which coexisitence could occur, and ATA effects
tended to be negative, weakening total storage effects.

%G3
Fulfilling goal G3 of the Introduction (to explore, using the lottery model, whether species
can coexist with ATAs, but not without, or vice versa), our results show that ATAs can, indeed, sometimes
make the difference between coexistence and exclusion: both ``ATA rescue'' and ``ATA
exclusion'' (see Methods) are possible.
For some of the panels on Fig. \ref{fig:fig3}, the IGR line falls between the ATA
line and the $x$ axis for some values of $\sigma$, e.g., see panel H. 
Such cases correspond to parameter combinations
for which the presence or absence of ATAs determine whether the species coexist or not. In the 
case of panel H, for values of $\sigma$ around $4$, 
species 1 is excluded with ATAs but the two species can coexist without ATAs. 
Fig. \ref{fig:fig4} highlights additional examples. Orange lines on Fig. \ref{fig:fig4}
indicate the IGR when the effects of ATAs are included, and blue lines are the IGR when
the effects of ATAs are neglected. These two lines frequently crossed the $x$ axis in 
different places, showing that both ATA rescue space and ATA exclusion space could be non-empty
subsets of our trait-difference space (see Methods). For the parameters we considered, 
both ATA rescue and exclusion were more common for larger values of $\sigma$.
ATA rescue appeared more likely for smaller $\delta$, whereas ATA exclusion appeared
more common for larger $\delta$. Although bands of ATA rescue or exlusion on Fig.
\ref{fig:fig4} appear narrow, they nevertheless do occur,

Althoug ATA rescue or exclusion
occurs only for the minority of parameter combinations considered, this does not mean
ATAs are unimportant for species coexistance and biodiversity maintenance - the same is
probably also true of

 Occurence of ATA rescue is slim and only occurs when $\delta=1$ and species traits are extremely similar or identical. Other $\delta$ values are more indicative of ATA exclusion events.  When $\delta<1$ ATA exlusion space increase in range and magnitude as $\sigma$ increases. There is indication that ATA exlcusion space also increases in magnitude as $\delta$ decreases, however changes in range have a more complex trend. Range increases across decreasing $\delta$ for all $\sigma$ values except $\sigma=6$. Here, the range is limited by an increasing $\Delta_i^{[E \nat C]}$ as trait space increases and the two IGR's are close to converging.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig3.pdf}
\caption{The contribution of ATAs to species coexistence is often meaningfully large compared to 
other components of the MTC decomposition. All panels show the IGR and the decomposition of
the IGR provided by the extended MTC, equation (\ref{eq:full_decomp}). For each panel, 
$\delta$ and $\mu_1-\mu_2$ 
are fixed at values specified in the margins, and the IGR and components of the decomposition
are plotted against $\sigma$. The contribution of ATAs (red line) frequently 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the IGR line (orange) falls 
betweed the ATA line (red)
and the $x$ axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of the weaker competitor. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3,5)}, so are not displayed. This plot uses $q_{ij}=1$. An 
analogous plot (SI Fig. \ref{SI-fig:fig3_alt}) was generated 
using $q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$. Though results differed
in detail, main conclusions were the same. \label{fig:fig3}}
\end{figure}
%***DAN: Dan has edited this caption

>>>>

%G2
Left- or right-tail associated noise $(b_1,b_2)$, with $b_i$ normally distributed with mean $\mu_i$ and standard deviation 
$\sigma$, was generated as pictured in Fig. \ref{fig:noise}, with details in SI section \ref{SI-sect:noise}.
To achieve goal G2 from the Introduction, i.e., to explore using a simple model the relative magnitudes of the contributions 
to coexistence that come from ATA versus other mechanisms, we plotted the terms $\Delta_i^0$, $\Delta_i^E$, $\Delta_i^C$, 
$\Delta_i^{(E\shp C)}$, $\Delta_i^{[E \nat C]}$ and $\Delta_i^{[EC]}$ [see (\ref{eq:full_decomp})] for the lottery model
against the independent parameters of the model, $\mu_1$, $\mu_2$, $\sigma$ and $\delta$; and then we compared their
values on the plots. Some mathematical shortcuts, summarized here, greatly simplified this plotting. First, we showed 
that, for the 
lottery model, none of the components in (\ref{eq:full_decomp}) depends independently on $\mu_1$ and $\mu_2$;
instead they all depend only on the difference $\mu_1-\mu_2$ (SI section \ref{SI-sect:math_simplifications}). So we only plotted 
against $\mu_1 - \mu_2$. Second, we showed that the values of all components in (\ref{eq:full_decomp}) were the same 
for both our left- and right-tail associated noises (SI section \ref{SI-sect:math_simplifications}). So we only generated plots 
for left-tail associated noise. For each combination of $\delta=0.2, 0.4, 0.6, 0.8$
and $\mu_1-\mu_2=0,-0.5,-2,-4$,
we plotted all the components of (\ref{eq:full_decomp}) against $\sigma$. 
Relative importance of contributions was judged from the 
plots to assess whether and to what extent ATA effects on coexistence were meaningfully large for the lottery model, compared to 
the contributions of other mechanisms.
Mathematical simplications for the $\delta=1$ case make it straightforward to see that
coexistence is impossible and that the case is relatively trivial (SI 
section \ref{SI-sect:delta1}). So that case is not considered further.  
The simulation methods of \cite{Ellner_2019} were used to compute the components of (\ref{eq:full_decomp}).
Computationally efficient means of carrying out these simulations, for the lottery model, are elaborated in 
SI section \ref{SI-sect:efficient_computation}.

%G3
To acheive goal G3 from the Introduction i.e., to explore, using the lottery model, the likelihood that species can coexist with
ATAs but not without, or vice versa, we plotted the IGR, $\olr_{i \bs i}=\olr_{i \bs i} -q_{ij}\olr_{j \bs i}$, 
as well as the
IGR minus its ATA contribution, $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$. 
For fixed values of $\sigma$ and $\delta$ and for left-tail associated noise, 
we were interested in the values of $\mu_1-\mu_2$ for which one of these
quantities is positive and the other is negative; this set of value characterizes the species which can coexist
with ATAs, but not without, or vice versa. The one dimensional space of values $\mu$ is a species trait space,
and $\mu_1$ and $\mu_2$ can be regarded as two points in it. But we instead define the ``trait-difference space'' of values $\mu_1-\mu_2$. 
Points of the trait-difference space characterize possible combinations of traits of the two competing species in our model. For fixed values of 
$\delta$ and $\sigma$, we defined the subset, $\psi_{\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of 
values $\mu_1 - \mu_2$ for which the the IGR, $\olr_{i \bs i}$, is positive. This 
is the subspace for which the two species in the lottery model coexist in the presence of ATAs.
We also defined the subset, $\psi_{\lnot\text{ATA}}(\delta,\sigma)$, of trait-difference space to be the set of values
$\mu_1 - \mu_2$ for which the metric $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$ is positive.
This is the subspace for which the two species in the lottery model coexist in the absence of 
ATAs. The symbol $\lnot$ denotes logical negation, so $\lnot \text{ATA}$ is intended to indicate the
absence of ATAs. We defined the ``ATA rescue space" associated with given values of $\delta$ and $\sigma$ to be the 
set difference $\psi_{\text{ATA}}(\delta,\sigma) \setminus \psi_{\lnot\text{ATA}}(\delta,\sigma)$, i.e., the
values $\mu_1 - \mu_2$ for which species coexist in the presence of ATAs, but not in their absence. 
Similarly, we define ``ATA exclusion space'' to be the set difference 
$\psi_{\lnot\text{ATA}}(\delta,\sigma) \setminus \psi_{\text{ATA}}(\delta,\sigma)$,
i.e., $\mu_1 - \mu_2$ values for which coexistence occurs in the absence, but not the presence, of ATAs.
On plots of $\olr_{i \bs i}$ and $\olr_{i \bs i} -\Delta_i^{[E \nat C]}$
against $\mu_1 - \mu_2$, we highlighted the ATA rescue and exclusion spaces. We plotted against
$\mu_1 - \mu_2$ for all combinations of $\delta=0.2, 0.4, 0.6, 0.8$ and $\sigma=2,4,5,6$.

@@

@@

%G2 
Fulfilling goal G2 of the Introduction (to explore, using the lottery model, the relative magnitudes 
of the contributions to coexistence that come from ATAs compared to other mechanisms), Fig.
\ref{fig:fig3} shows that the ATA term of (\ref{eq:full_decomp}) sometimes, but not always, has 
magnitude comparable to the magnitudes of other terms of the extended MTC decomposition. 
Magnitudes of $\Delta_i^{[E \nat C]}$ tended to be larger, relative to the magnitudes of the 
other terms of (\ref{eq:full_decomp}), for smaller-magnitude differences between $\mu_1$ and 
$\mu_2$, i.e., for closer-to-equal competitors. 
In the $\mu_1=\mu_2$ case (panels A-D), for which the two species were equal 
competitors, storage effects (blue plus red
lines on Fig. \ref{fig:fig3}) were the only means by which coexisitence could occur, and ATA effects
tended to be negative, weakening total storage effects.

%G3
Fulfilling goal G3 of the Introduction (to explore, using the lottery model, whether species
can coexist with ATAs, but not without, or vice versa), our results show that ATAs can, indeed, sometimes
make the difference between coexistence and exclusion; both ``ATA rescue'' and ``ATA
exclusion'' (see Methods) are possible.
For some of the panels on Fig. \ref{fig:fig3}, the IGR line falls between the ATA
line and the $x$ axis for some values of $\sigma$, e.g., see panel H. 
Such cases correspond to parameter combinations
for which the presence or absence of ATAs determine whether the species coexist or not. In the 
case of panel H, for values of $\sigma$ around $4$, 
species 1 is excluded with ATAs but the two species can coexist without ATAs. 
Fig. \ref{fig:fig4} highlights additional examples. Orange lines on Fig. \ref{fig:fig4}
represent the IGR when the effects of ATAs are included, and blue lines are the IGR when
the effects of ATAs are removed. These two lines frequently crossed the $x$ axis in 
different places, showing that both ATA rescue space and ATA exclusion space can be non-empty
subsets of our trait-difference space (see Methods). For the parameters we considered, 
both ATA rescue and exclusion were more common for larger values of $\sigma$.
ATA rescue appeared more likely for smaller $\delta$, whereas ATA exclusion appeared
more common for larger $\delta$. Although bands of ATA rescue or exlusion on Fig.
\ref{fig:fig4} appear narrow, they nevertheless do occur, which means ATAs can mean the 
difference between coexistence and extirpation of one species in some cases.

@@

\begin{figure}
\includegraphics[width=\textwidth]{../results_figs/fig3.pdf}
\caption{The contribution of ATAs to species coexistence is often meaningfully large compared to 
other components of the MTC decomposition. All panels show the IGR and the decomposition of
the IGR provided by the extended MTC, equation (\ref{eq:full_decomp}). For each panel, 
$\delta$ and $\mu_1-\mu_2$ 
are fixed at values specified in the margins, and the IGR and components of the decomposition
are plotted against $\sigma$. The contribution of ATAs (red line) frequently 
has magnitude comparable to other components, indicating that ATAs can contribute meaningfully
to species coexistence or failure to coexist. In some instances, the IGR line (orange) falls 
betweed the ATA line (red)
and the $x$ axis, indicating that, for those parameter values, ATAs make the difference between species
coexistence and exclusion of the weaker competitor. Standard errors of plotted quantities 
were never greater than \Sexpr{round(se3,5)}, so are not displayed. This plot uses $q_{ij}=1$. An 
analogous plot (SI Fig. \ref{SI-fig:fig3_alt}) was generated 
using $q_{ij}=\exp(\mu_i)/[(1-\delta)\exp(\mu_j)+\delta\exp(\mu_i)]$. Though results differed
in some details, main conclusions were the same. \label{fig:fig3}}
\end{figure}
%***DAN: Dan has edited this caption
